---
title: "Easy to Use and Share Data Labels in R with Dictionaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Easy to Use and Share Data Labels in R with Dictionaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(labeler)
```

## The Problem with Labelled Data in R

Historically, software like SPSS, SAS, and Stata have worked seamlessly with labelled data. In R, the closest equivalent is factors, which actually work in reverse—assigning numeric codes to text values.

The R ecosystem has several packages for labelled data: `haven`, `labelled`, and `sjlabelled`. However, they all share common limitations:

| Limitation | Impact |
|------------|--------|
| **Labels tied to files** | Must read from SPSS/Stata files or manually re-assign labels |
| **Lost on operations** | Many R operations strip label attributes |
| **No sharing mechanism** | Can't easily share label definitions across projects or teams |
| **No database support** | Labels live only in memory or specific file formats |
| **R-only** | No interoperability with Python or other languages |

## Why labeler?

**`labeler` solves these problems with a dictionary-based approach:**

```{r, echo=FALSE}
comparison <- data.frame(
  Feature = c(
    "Shareable label definitions",
    "Survives R operations",
    "JSON export/import",
    "Database storage",
    "R/Python interoperability",
    "Auto-generate from data"
  ),
  haven = c("❌", "❌", "❌", "❌", "❌", "❌"),
  labelled = c("❌", "❌", "❌", "❌", "❌", "✅"),
  sjlabelled = c("❌", "❌", "❌", "❌", "❌", "❌"),
  labeler = c("✅", "✅", "✅", "✅", "✅", "✅")
)
knitr::kable(comparison, align = "lcccc")
```

> **Key insight:** Labels are defined once in a `Dict` object, stored externally (JSON or database), and applied on-demand. Lost labels? Just re-apply the dictionary.

---

## Creating a Dictionary

A `Dict` is created with the `Dict()` function. Each variable definition includes:

- **label:** Description (optionally with short name before "||")
- **labels:** Named vector for value labels
- **dtype:** (Optional) Target data type

```{r}
dict <- Dict(
  GENDER = list(
    label = "Gender || Gender of the person",
    labels = c("Man" = 1, "Woman" = 2),
    dtype = "integer"
  ),
  AGE = list(
    label = "Age || Age of the person",
    dtype = "integer"
  ),
  Married = list(
    label = "Marital Status || Is the person married?",
    labels = c("No" = 0, "Yes" = 1)
  ),
  metadata = list(name = "survey_data")
)

dict
```

### Auto-generate from labelled data

If you already have labelled data, generate a dictionary automatically:

```{r}
lbl_df <- set_Dict(lbl_df, lbl_df_dict)
generated_dict <- gen_Dict(lbl_df)
str(generated_dict)
```

### Browse interactively

```{r eval=FALSE}
dict_browser(dict)
```

---

## Applying Labels

Use `set_Dict()` to apply labels to a data frame:

```{r}
mydata <- data.frame(
  GENDER = c(rep(1, 5), rep(2, 5)),
  AGE = seq(1, 30, 3),
  Married = c(rep(0, 4), rep(1, 6))
)

labeled_data <- set_Dict(mydata, dict)
str(labeled_data)
```

Apply to specific variables only:

```{r}
partial <- set_Dict(mydata, dict, subset = c("GENDER"))
str(partial)
```

---

## Using Labels

`with_Dict()` converts codes to labels and/or renames columns:

```{r}
# Values → Labels
with_Dict(mydata, dict, use_labels = TRUE, use_label = FALSE)
```

```{r}
# Column names → Variable labels
with_Dict(mydata, dict, use_label = TRUE, use_labels = FALSE)
```

Works great with dplyr:

```{r}
mydata %>%
  set_Dict(dict, label_type = "name") %>%
  dplyr::group_by(GENDER, Married) %>%
  dplyr::count() %>%
  with_Dict(use_labels = TRUE)
```

---

## Sharing Dictionaries: JSON

Export your dictionary as JSON—readable by R, Python, JavaScript, or any language:

```{r}
json_str <- to_json(dict)
cat(json_str)
```

Save to file:

```{r eval=FALSE}
to_json(dict, path = "survey_dict.json")
```
  
Import from file:

```{r eval=FALSE}
my_dict <- from_json("survey_dict.json")
```

### JSON Schema

The JSON follows a versioned schema for future compatibility:

```json
{
  "$schema": "https://labeler.adatar.do/dict-schema/v1",
  "metadata": { "name": "survey_data" },
  "variables": {
    "GENDER": { "label": "...", "labels": {...} }
  }
}
```

---

## Storing Dictionaries: Database

Store dictionaries in **any SQL database** via DBI. This enables:

- **Team collaboration** — Single source of truth
- **Version control alternative** — Centralized dictionary management
- **Cross-project reuse** — One dictionary, many projects
- **R/Python sharing** — Both languages can read the same table

```{r, eval=FALSE}
library(DBI)

# Connect to database (PostgreSQL, MySQL, SQLite, SQL Server, etc.)
con <- dbConnect(RSQLite::SQLite(), "my_project.sqlite")

# Save dictionary (table auto-created on first use)
db_save_dict(con, dict)

# List all dictionaries
db_list_dicts(con)

# Load a dictionary
my_dict <- db_load_dict(con, "survey_data")

# Update existing dictionary
db_save_dict(con, updated_dict, overwrite = TRUE)

# Delete
db_delete_dict(con, "old_dict")

dbDisconnect(con)
```

The `labeler_dicts` table uses simple TEXT columns for maximum compatibility across database systems.

---

## Summary

| Task | Function |
|------|----------|
| Create dictionary | `Dict()` |
| Generate from data | `gen_Dict()` |
| Apply to data | `set_Dict()` |
| Use labels | `with_Dict()` |
| Export to JSON | `to_json()` |
| Import from JSON | `from_json()` |
| Save to database | `db_save_dict()` |
| Load from database | `db_load_dict()` |

**labeler** makes labelled data in R truly portable, shareable, and interoperable.